Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-12 analysis/paper/paper.Rmd
minor edits

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Chen2015; @Hao2022; @Weisskopf2015]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that both millet and rice have been introduced into southwestern China at the latest in 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].
